Validating risk scores

### 2020-10-09 manuscript/risk_scores.Rmd
Further work on risk scores

```{r}
baseline <- data$patient_data %>%
  mutate(alive_12_admission = case_when(
    last_record >= 12 ~ TRUE,
    outcome == "Discharged" ~ TRUE,
    outcome == "Death" ~ FALSE,
    TRUE ~ NA),
    alive_30_admission = case_when(
    last_record >= 30 ~ TRUE,
    outcome == "Discharged" ~ TRUE,
    outcome == "Death" ~ FALSE,
    TRUE ~ NA)
  )

@@

# acp_data %>%
#   group_by(ACP_grade, worst_breathing_12) %>%
#   summarise(mean(alive_12))
```
```{r}
li_et_al_data <- data.frame(
  ACP_grade = c(rep(1, 194), rep(2, 145), rep(3, 95)),
  alive_12 = c(rep(TRUE, 194), #Grade 1, 0% mortality
               rep(FALSE, 8), rep(TRUE, 145-8), #Gred 2, 5.6% mortality
               rep(FALSE, 32), rep(TRUE, 95 - 32) #Grade 3, 33.2% mortality
  ))

@@

```{r}
outcome_caption = c("alive_12" = "12-day mortality", "alive_30" = "30-day mortality")
score_caption = c("ACP_grade" = "Li et al. (ACP)")

my_auc <- function(f, d) {
  roc = pROC::roc(f, d, direction = "<", levels = c(TRUE, FALSE))
  ci = pROC::ci.auc(roc)

  outcome = as.character(formula.tools::lhs(f))
  score = as.character(formula.tools::rhs(f))

  list(roc = roc,
       summary = data.frame(
         auc = ci[2],
         auc_low = ci[1],
         auc_high = ci[3],
         outcome = outcome,
         score = score
       )
  )
}

plot_my_auc <- function(my_auc) {
  caption = paste0(score_caption[my_auc$summary$score], ", ", outcome_caption[my_auc$summary$outcome])
  plot(my_auc$roc, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE, main = caption)
}

plot_all_auc <- function(my_auc_list) {
  my_auc_list %>% walk(plot_my_auc)
}
```

@@

@@

```{r}
chen_liu_data <- data$marker_data_wide %>%  
  mutate(has_all_values = !is.na(CRP) & !is.na(d_dimer) & !is.na(lymphocyte_count),
         d_dimer_converted = d_dimer * 2 / 1000) %>%
  group_by(patient_id) %>%
  filter(any(has_all_values)) %>%
  summarise(
    first_day = min(day[has_all_values]),
    first_CRP = CRP[day == first_day],
    first_d_dimer_converted = d_dimer_converted[day == first_day],
    first_lymphocyte_count = lymphocyte_count[day == first_day],
    breathing_first = breathing[day == first_day],
    .groups = "drop") %>% 
  inner_join(data$patient_data, by = c("patient_id")) %>%
  mutate(    
    inclusion_criterium = breathing_first < "NIPPV" & (outcome %in% c("Discharged", "Death") | last_record >= first_day + 7),
    score =  -4.21167 + 0.058699 * age + 0.013391 * first_CRP + 0.112055 * first_d_dimer_converted - 1.983733 * first_lymphocyte_count,
    risk = 1 / (1 + exp(-score)),
    risk_lower = 1 / (1 + exp(-score + 1.96*0.1721)),
    risk_upper = 1 / (1 + exp(-score - 1.96*0.1721)),
    alive_12 = case_when(last_record >= first_day + 12 | outcome == "Discharged" ~ TRUE,
                         outcome == "Death" ~ FALSE,
                         TRUE ~ NA),
    alive_30 = case_when(last_record >= first_day + 30 | outcome == "Discharged" ~ TRUE,
                         outcome == "Death" ~ FALSE,
                         TRUE ~ NA),
    alive_all = case_when(outcome == "Discharged" ~ TRUE,
                         outcome == "Death" ~ FALSE,
                         TRUE ~ NA),
    alive_imputed = outcome != "Death"

  ) %>% select(age, starts_with("first"), score, starts_with("risk"), breathing_first, starts_with("alive"), inclusion_criterium)
```

```{r}
chen_liu_data %>% ggplot(aes(x = score, y = alive_30)) + geom_point()
```

```{r}
roc_chen_liu_12 <- pROC::roc(alive_12 ~ score, chen_liu_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
roc_chen_liu_30 <- pROC::roc(alive_30 ~ score, chen_liu_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
roc_chen_liu_all <- pROC::roc(alive_all ~ score, chen_liu_data, direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)

roc_chen_liu_12_incl <- pROC::roc(alive_12 ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
roc_chen_liu_30_incl <- pROC::roc(alive_30 ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
roc_chen_liu_all_incl <- pROC::roc(alive_all ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)
roc_chen_liu_imputed_incl <- pROC::roc(alive_imputed ~ score, chen_liu_data %>% filter(inclusion_criterium), direction = "<", levels = c(TRUE, FALSE), plot = TRUE, auc.polygon=TRUE, max.auc.polygon=TRUE, print.auc = TRUE)

@@

Not sure what "severe" means

@@

bello_chavolla_data <- baseline %>%
  inner_join(atb_first_day, by = "patient_id") %>%
  mutate(bc_base =  3 *  (age >= 65) +
           diabetes +
           5 * (diabetes & age < 40) +
           (-6) * (age < 40) +
           obesity +
           3 * renal_disease +
           COPD,
         bc0 = bc_base + 7 * atb0,
         bc1 = bc_base + 7 * atb1
  )

>>>>

```{r}
at_admission <- data$patient_data %>%
  mutate(alive_12 = case_when(
    last_record >= 12 ~ TRUE,
    outcome == "Discharged" ~ TRUE,
    outcome == "Death" ~ FALSE,
    TRUE ~ NA),
    alive_30 = case_when(
    last_record >= 30 ~ TRUE,
    outcome == "Discharged" ~ TRUE,
    outcome == "Death" ~ FALSE,
    TRUE ~ NA)
  )

@@

# acp_data %>%
#   group_by(ACP_grade, worst_breathing_12) %>%
#   summarise(mean(alive_12))
```

```{r}
li_et_al_data <- data.frame(
  ACP_grade = c(rep(1, 194), rep(2, 145), rep(3, 95)),
  alive_12 = c(rep(TRUE, 194), #Grade 1, 0% mortality
               rep(FALSE, 8), rep(TRUE, 145-8), #Gred 2, 5.6% mortality
               rep(FALSE, 32), rep(TRUE, 95 - 32) #Grade 3, 33.2% mortality
  ))

@@

@@

```{r}
compare_auc_to_orig(list(acp_auc_orig), list(acp_auc))
#acp_auc
```

@@

```{r}
chen_liu_data <- data$marker_data_wide %>%  
  mutate(has_all_values = !is.na(CRP) & !is.na(d_dimer) & !is.na(lymphocyte_count),
         d_dimer_converted = d_dimer * 2 / 1000) %>%
  group_by(patient_id) %>%
  filter(any(has_all_values)) %>%
  summarise(
    first_day = min(day[has_all_values]),
    first_CRP = CRP[day == first_day],
    first_d_dimer_converted = d_dimer_converted[day == first_day],
    first_lymphocyte_count = lymphocyte_count[day == first_day],
    breathing_first = breathing[day == first_day],
    .groups = "drop") %>% 
  inner_join(data$patient_data, by = c("patient_id")) %>%
  mutate(    
    inclusion_criterium = breathing_first < "NIPPV" & (outcome %in% c("Discharged", "Death") | last_record >= first_day + 7),
    chen_liu = -4.21167 + 0.058699 * age + 0.013391 * first_CRP + 0.112055 * first_d_dimer_converted - 1.983733 * first_lymphocyte_count,
    risk = 1 / (1 + exp(-chen_liu)),
    risk_lower = 1 / (1 + exp(-chen_liu + 1.96*0.1721)),
    risk_upper = 1 / (1 + exp(-chen_liu - 1.96*0.1721)),
    alive_12 = case_when(last_record >= first_day + 12 | outcome == "Discharged" ~ TRUE,
                         outcome == "Death" ~ FALSE,
                         TRUE ~ NA),
    alive_30 = case_when(last_record >= first_day + 30 | outcome == "Discharged" ~ TRUE,
                         outcome == "Death" ~ FALSE,
                         TRUE ~ NA),
    alive_all = case_when(outcome == "Discharged" ~ TRUE,
                         outcome == "Death" ~ FALSE,
                         TRUE ~ NA),
    alive_imputed = outcome != "Death"

  ) %>% select(age, starts_with("first"), chen_liu, starts_with("risk"), breathing_first, starts_with("alive"), inclusion_criterium)
```

```{r}
chen_liu_data %>% ggplot(aes(x = chen_liu, y = alive_30)) + geom_point()
```

```{r}
chen_liu_data_inclusion <- chen_liu_data %>% filter(inclusion_criterium)
chen_liu_auc <- list(
  my_auc(alive_12 ~ chen_liu, chen_liu_data),
  my_auc(alive_30 ~ chen_liu, chen_liu_data),
  my_auc(alive_all ~ chen_liu, chen_liu_data),
  my_auc(alive_imputed ~ chen_liu, chen_liu_data),
  my_auc(alive_12 ~ chen_liu, chen_liu_data_inclusion, "incl"),
  my_auc(alive_30 ~ chen_liu, chen_liu_data_inclusion, "incl"),
  my_auc(alive_all ~ chen_liu, chen_liu_data_inclusion, "incl"),
  my_auc(alive_imputed ~ chen_liu, chen_liu_data_inclusion, "incl")
)

plot_all_auc(chen_liu_auc)

```{r}
chen_liu_auc_orig <- my_auc_from_estimate(0.881, "alive", "chen_liu", "incl")
compare_auc_to_orig(list(chen_liu_auc_orig), list(chen_liu_auc))

@@

Not sure what "severe" means, will test both "not ventilated", "not needing Oxygen"

```{r}
shi_et_al_data <- data$marker_data_wide %>%  
  group_by(patient_id) %>%
  summarise(
    breathing_admission = max(breathing[!is.na(breathing) & day < 2]),
    not_ventilated_admission = breathing_admission < "NIPPV",
    not_oxygen_admission = breathing_admission < "Oxygen",
    .groups = "drop") %>% 
  inner_join(at_admission, by = c("patient_id")) %>%
  mutate(shi = (age >= 50) + (sex == "M") + has_hypertension_drugs,
         not_ventilated_all = case_when(
           outcome == "Death" ~ FALSE,
           outcome != "Discharged" ~ NA,
           worst_breathing < "NIPPV" ~ TRUE
         ),
         not_oxygen_all = case_when(
           outcome == "Death" ~ FALSE,
           outcome != "Discharged" ~ NA,
           worst_breathing < "Oxygen" ~ TRUE
         )
         )

shi_et_al_auc <- list(
  my_auc(alive_12 ~ shi, shi_et_al_data),
  my_auc(alive_30 ~ shi, shi_et_al_data),
  my_auc(not_ventilated_all ~ shi, shi_et_al_data),
  my_auc(not_oxygen_all ~ shi, shi_et_al_data),
  my_auc(not_ventilated_admission ~ shi, shi_et_al_data),
  my_auc(not_oxygen_admission ~ shi, shi_et_al_data)
)

plot_all_auc(shi_et_al_auc)
```

From Fig1 B,C:
```{r}
shi_et_al_reported <- data.frame(
  score = c(0,1,2,3), shi_severe_admission = c(0,0.057,0.19,0.4), shi_severe_all = c(0.083, 0.138,0.389,0.429)
)

shi_et_al_compare <- shi_et_al_data %>%
  group_by(shi) %>%
  summarise(
    across(
      one_of(c("not_ventilated_admission", "not_ventilated_all", "not_oxygen_admission", "not_oxygen_all")), 
      ~ mean(.x, na.rm = TRUE), .names = "our_{col}"), 
    n_our_admission = sum(!is.na(not_ventilated_admission)), 
    n_our_all = sum(!is.na(not_ventilated_all)),
    .groups = "drop") %>%
  inner_join(shi_et_al_reported, by = c("shi" = "score"))

shi_et_al_compare %>%
  transmute(score = shi, `Shi et al. (admission)` = shi_severe_admission, `Ventilated (admission)` = 1 - our_not_ventilated_admission, `Oxygen (admission)` = 1 - our_not_oxygen_admission, N = n_our_admission)

shi_et_al_compare %>%
  transmute(score = shi, `Shi et al. (hospitalization)` = shi_severe_all, `Ventilated (hospitalization)` = 1 - our_not_ventilated_all, `Oxygen (hospitalization)` = 1 - our_not_oxygen_all, N = n_our_all)

@@

bello_chavolla_data <- at_admission %>%
  inner_join(atb_first_day, by = "patient_id") %>%
  mutate(bc_base =  3 *  (age >= 65) +
           diabetes +
           5 * (diabetes & age < 40) +
           (-6) * (age < 40) +
           obesity +
           3 * renal_disease +
           COPD,
         bc0 = bc_base + 7 * atb0,
         bc1 = bc_base + 7 * atb1
  )
